A comparison of model-based and design-based approaches for spatial data.

### 2021-06-08 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
simulation outline discussion points.
some introduction edits

# Introduction

@@

@@

@de1990model give an early comparison of the paradigms for spatial data, quashing the belief that classical sampling-based methods could not be used for spatially correlated data. Thereafter, several comparisons between model-based and design-based for spatial data have been considered, but they tend to compare spatial models to IRS (Independent Random Sampling) or SIRS (stratified IRS) [@brus1997random]. No comparison has been made between spatial models and spatially balanced design-based approaches.

@@

* hansen1983evaluation - general overview (in non spatial context), talk about population vs superpopulation give a set of guiding principles in the discussion

* verhoef2002sampling - strictly defines classical sampling, compares FPBK to classical, BK to classical

* wang2012spatial - sampling and inference steps, thorough overview in terms of spatial context, lots of references

# Background

@@

# Numerical Study

@@

# Discussion

>>>>

# Introduction {#sec:introduction}

@@

<!-- Brief introduction to model based and design based -->

There are two approaches to statistical inference: design-based and model-based. Both approaches aim to study a population. Because data generally cannot be sampled at all units in the population (known as population units), the population units must be sampled. In the design-based approach, the samples are often used to estimate parameters of a population. Properties of these estimators are derived through random selection of the samples. In the model-based approach, the samples are often used to predict random variables of a population. Properties of these predictions are derived through a set of distributional assumptions that specify the population's structure. Randomness is crucial to both approaches but for different reasons; the design-based approach uses randomness when selecting a sample of population units, and the model-based approach uses randomness when specifying distributional assumptions. @sterba2009alternative provides historical context for each approach, and @hansen1983evaluation discusses the advantages of each approach [@hansen1983evaluation].

<!-- Shift focus to spatial data -->

Though the design-based and model-based approaches apply to statistical inference in a broad sense, we focus comparing these approaches for spatial data. @de1990model gave an early comparison of these approaches, quashing the belief that design-based approaches could not be used for spatially correlated data. Thereafter, several comparisons between design-based and model-based for spatial data have been considered, but they tend to compare design-based approaches that ignore spatial location in sampling to model-based approaches [@brus1997random, @verhoef2002sampling, @verhoef2008spatial]. No comparison has been made design-based approaches that incorporate spatial location in sampling and model-based approaches.

@@

@@

wang2012spatial - sampling and inference steps, thorough overview in terms of spatial context, lots of references

<!-- Outline for the rest of the paper -->

In Section \ref{sec:background}, we compare sampling and estimation procedures between the design-based approach and the model-based approach. In Section \ref{sec:numstudy}, we use simulated and real data to compare the properties of parameter estimates from both aproaches. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

# Background {#sec:background}

@@

# Numerical Study {#sec:numstudy}

cases

__Base Simulations__

* both good: correctly specified model with high correlation
* break model: non-gaussian errors
* break design: small area estimation

* both good?: misspecified covariance model with high correlation
* break both? non-gaussian areas with smaller sample size

* change n or sampling fraction

* model-based: how should sample be drawn?

@@

# Discussion {#sec:discussion}

### 2021-06-08 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.tex
some introduction edits

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

Several comparisons between model-based and design-based for spatial
data have been considered, but they tend to compare spatial models to
IRS (Independent Random Sampling) or SIRS (stratified IRS). No
comparison has been made between spatial models and spatially balanced
design-based approaches.

\hypertarget{background}{%
\section{Background}\label{background}}

@@

The design-based approach to inference uses characteristics of the
sampling design to estimate parameters of interest. The design-based
approach assumes the data, \(\mathbf{Y}\), are fixed. Randomness is
incorporated through the random selection of sampling units based on
probabilities derived using the sampling design itself. Some examples of
commonly used sampling designs include independent random sampling
(IRS), stratified random sampling, and cluster sampling. Viewing the
data as fixed and incorporating randomness through the sampling design
yields estimators having very few assumptions. Asymptotic properties of
these types of estimators are derived using limiting arguments. Means
and totals, for example, are asymptotically Gaussian. Särndal et al.
(2003) and Lohr (2009) provide thorough reviews of the design-based
approach.

@@

@@

Design-Based and Model-Based Comparisons

There have been many comparisons between the two paradigms\ldots\ldots.

@@

The Horvitz-Thompson estimator (Horvitz and Thompson, 1952) and its
continuous analog (Cordy, 1993) yield unbiased estimates of population
totals (and means). If \(\tau\) is a population total, then the
Horvitz-Thompson estimator of \(\tau\), denoted by \(\hat{\tau}_{HT}\),
is given by \begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Y_i \pi_i^{-1},
\end{align} where \(Y_i\) is the observed value of the \(i\)th sampling
unit, and \(\pi_i\) is the inclusion probability of the \(i\)th sampling
unit. Estimating the variance of \(\hat{\tau}_{HT}\) directly relies on
knowing the probability that \(s_i\) and \(s_j\) are both included in
the sample (second-order inclusion probability), which can be unfeasible
to compute. Furthermore, this variance estimator does not incorporate
spatial locations. To address these challenges, (Stevens Jr and Olsen,
2003) proposed an alternative variance estimator: the local neighborhood
variance estimator. The local neighborhood variance estimator does not
require second-order inclusion probabilities. It also incorporates
spatial locations, which tends to reduce the estimated variance of
compared \(\hat{\tau}_{HT}\) to a variance estimator ignoring spatial
location (Stevens Jr and Olsen, 2003).

@@

@@

@@

@@

@@

@@

@@

>>>>

\hypertarget{sec:introduction}{%
\section{Introduction}\label{sec:introduction}}

@@

There are two approaches to statistical inference for spatial data:
design-based and model-based. In the design-based approach, statistical
inference is based on random sampling. In the model-based approach,
statistical inference is based on a set of distributional assumptions
for the data.

In either approach, statistical inference requires estimating parameters
of interest. Often, these parameters are means or totals. In the
design-based approach, parameters correspond to some population. In the
model-based approach, parameters correspond to some statistical process,
sometimes called a superpopulation.

Spatial data paragraph

define sampling design, define parameters, define model-based, sample,
population, super-population?, etc.

introduction of spatial data / definition of spatial and maybe general
background?

give some examples of possible spatial data

2-3 sentences about each design

goals and outline of paper

De Gruijter and Ter Braak (1990) give an early comparison of the
paradigms for spatial data, quashing the belief that classical
sampling-based methods could not be used for spatially correlated data.
Thereafter, several comparisons between model-based and design-based for
spatial data have been considered, but they tend to compare spatial
models to IRS (Independent Random Sampling) or SIRS (stratified IRS)
(Brus and De Gruijter, 1997). No comparison has been made between
spatial models and spatially balanced design-based approaches.

brus2020statistical: most similar to what this paper will be, directly
comparing the two approaches. This has a lot of other potentially useful
references to go back to. (published in 2021 so likely doesn't have many
that have cited it yet).

\begin{itemize}
\item
  hansen1983evaluation - general overview (in non spatial context), talk
  about population vs superpopulation give a set of guiding principles
  in the discussion
\item
  verhoef2002sampling - strictly defines classical sampling, compares
  FPBK to classical, BK to classical
\item
  wang2012spatial - sampling and inference steps, thorough overview in
  terms of spatial context, lots of references
\end{itemize}

In Section \ref{sec:background}

\hypertarget{sec:background}{%
\section{Background}\label{sec:background}}

@@

The design-based approach to inference uses characteristics of the
sampling design to estimate parameters of interest. The design-based
approach assumes the data, \(\mathbf{Z}\), are fixed. Randomness is
incorporated through the random selection of sampling units based on
probabilities derived using the sampling design itself. Some examples of
commonly used sampling designs include independent random sampling
(IRS), stratified random sampling, and cluster sampling. Viewing the
data as fixed and incorporating randomness through the sampling design
yields estimators having very few assumptions. Asymptotic properties of
these types of estimators are derived using limiting arguments. Means
and totals, for example, are asymptotically Gaussian. Särndal et al.
(2003) and Lohr (2009) provide thorough reviews of the design-based
approach.

@@

We only use FPBK in this paper in order to focus more on comparing the
design-based and model-based approaches. However, k-nearest-neighbors
(Fix and Hodges, 1951; Ver Hoef and Temesgen, 2013), random forest
(Breiman, 2001), Bayesian models (Chan-Golston et al., 2020), among
others, can also be used to obtain predictions for a mean or total from
spatially correlated responses in a finite population setting.

@@

Several comparisons between the two paradigms have been made in many
different contexts. Sterba (2009) give some history of the two paradigms
as well as some applications in psychological contexts. Cooper (2006)
review the two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. We note that while there has been substantial research and
development into estimators that use both design and model-based
principles (see e.g.~Cicchitelli and Montanari (2012), Chan-Golston et
al. (2020) for a Bayesian approach, CITE some more of these), our goal
is not to expand upon or improve these methods.

@@

The Horvitz-Thompson estimator (Horvitz and Thompson, 1952) and its
continuous analog (Cordy, 1993) yield unbiased estimates of population
totals (and means). If \(\tau\) is a population total, then the
Horvitz-Thompson estimator of \(\tau\), denoted by \(\hat{\tau}_{HT}\),
is given by \begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the observed value of the \(i\)th sampling
unit, and \(\pi_i\) is the inclusion probability of the \(i\)th sampling
unit. Estimating the variance of \(\hat{\tau}_{HT}\) directly relies on
knowing the probability that \(s_i\) and \(s_j\) are both included in
the sample (second-order inclusion probability), which can be unfeasible
to compute. Furthermore, this variance estimator does not incorporate
spatial locations. To address these challenges, (Stevens Jr and Olsen,
2003) proposed an alternative variance estimator: the local neighborhood
variance estimator. The local neighborhood variance estimator does not
require second-order inclusion probabilities. It also incorporates
spatial locations, which tends to reduce the estimated variance of
compared \(\hat{\tau}_{HT}\) to a variance estimator ignoring spatial
location (Stevens Jr and Olsen, 2003).

@@

cases

\textbf{Base Simulations}

\begin{itemize}
\item
  both good: correctly specified model with high correlation
\item
  break model: non-gaussian errors
\item
  break design: small area estimation
\item
  both good?: misspecified covariance model with high correlation
\item
  break both? non-gaussian areas with smaller sample size
\item
  change n or sampling fraction
\item
  model-based: how should sample be drawn?
\end{itemize}

@@

\leavevmode\hypertarget{ref-breiman2001random}{}%
Breiman, L., 2001. Random forests. Machine learning 45, 5--32.

\leavevmode\hypertarget{ref-brus1997random}{}%
Brus, D., De Gruijter, J., 1997. Random sampling or geostatistical
modelling? Choosing between design-based and model-based sampling
strategies for soil (with discussion). Geoderma 80, 1--44.

\leavevmode\hypertarget{ref-chan2020bayesian}{}%
Chan-Golston, A.M., Banerjee, S., Handcock, M.S., 2020. Bayesian
inference for finite populations under spatial process settings.
Environmetrics 31, e2606.

@@

\leavevmode\hypertarget{ref-cicchitelli2012model}{}%
Cicchitelli, G., Montanari, G.E., 2012. Model-assisted estimation of a
spatial population mean. International Statistical Review 80, 111--126.

\leavevmode\hypertarget{ref-cooper2006sampling}{}%
Cooper, C., 2006. Sampling and variance estimation on continuous
domains. Environmetrics: The official journal of the International
Environmetrics Society 17, 539--553.

@@

\leavevmode\hypertarget{ref-de1990model}{}%
De Gruijter, J., Ter Braak, C., 1990. Model-free estimation from spatial
samples: A reappraisal of classical sampling theory. Mathematical
geology 22, 407--415.

@@

\leavevmode\hypertarget{ref-fix1951discriminatory}{}%
Fix, E., Hodges, J.L., 1951. Discriminatory analysis, nonparametric
discrimination: Consistency properties. USAF School of Aviation
Medicine.

@@

\leavevmode\hypertarget{ref-sterba2009alternative}{}%
Sterba, S.K., 2009. Alternative model-based and design-based frameworks
for inference from samples to populations: From polarization to
integration. Multivariate behavioral research 44, 711--740.

@@

\leavevmode\hypertarget{ref-ver2013comparison}{}%
Ver Hoef, J.M., Temesgen, H., 2013. A comparison of the spatial linear
model to nearest neighbor (k-nn) methods for forestry applications. PloS
one 8, e59129.
